# Application to exposure dating

```{r include=FALSE}
library("TCNtools")
```

## Back to the evolution of concentration

We first consider the evolution of concentration with time $t$.
The computation will be carried out at the surface ($z=0$), but this could be done at any arbitrary depth.
We also consider that $\varepsilon = 0$ and that there is no inheritance.
The starting equation above becomes, $$
C(t)=\sum_i \frac{P_i}{\lambda}(1-e^{-\lambda t})
$$ 
For the sake of the example we are going to compute that by hand, and compare with our results above obtained with the `solv_conc_eul` function,

```{r `}
# set up
altitude = 1000 # elevation in m
latitude = 45 # latitude in degrees
nuc = "Be10"
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
t =  seq(0,200e3,by=100) # a vector containing time from 0 to 100 ka by 100 a steps
Pspal = prm[1,nuc]*S$Nneutrons # scaled spallation production rate in at/g/y (st scaling)
Pstop = prm[2,nuc]*S$Nmuons # scaled stopped muons  production rate in at/g/y
Pfast = prm[3,nuc]*S$Nmuons # scaled fast muons  production rate in at/g/y
lambda = prm[4,nuc] # radioactive decay (1/y)
C = (Pspal + Pstop + Pfast) / lambda * (1-exp(-lambda*t))
plot(t/1000,C,type="l",col="coral",lwd=3,ylab="Concentration (at/g)",xlab="Time (ka)")
grid()
# check if this is ok
C2 = solv_conc_eul(0,0,t,0,prm[,nuc],S,Lambda) # compute concentration for pure exposure
lines(t/1000,C2,lty=2)

```

On thing to note in this graph is that time $t$ is expressed as *time elapsed since start of exposure*, which we can not directly translate into the age BP, when dealing with radioactive decay and time variations in production rate. 

## Interpretation in terms of exposure ages
 We are no going to make some simple calculations to convert our concentrations in terms of exposure age. 
It basically solving for $t$ the equation $C_{mod}(t) = C_{mes}$. The left hand side is our modelled concentration, obtained for example with ou simple Eulerian model (`solv_conc_eul`), where we need to keep in mind all the hypotheses we make ($z=0$, $\varepsilon=0$, $C_0=0$) 

We are going to use an example from @protin2019climatic, using sample *ARG-16-9*. Here are the characteristics of this sample.

```{r}
rho = 2.7
altitude = 2252 # elevation in m
latitude = 45.97 # latitude in degrees
longitude = 6.96 # longitude in degrees
Ss = 0.939*exp(-2.6*rho/Lambda[1]) # factor accounting for topographic shielding and sample thickness
Cmes = 26.8e4 # measured concentration (at/g)
Cmes_e = 1.3e4
#
P = atm_pressure(alt=altitude,lat=latitude,lon=longitude,model="era40") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

Now, for example we can make some guess about the exposure age and see what is the difference with the measured concentration.

```{r}
age = 10e3 # a 
Cmod = solv_conc_eul(0,0,age,0,prm[,"Be10"],S*Ss,Lambda) # compute concentration
```

We could try make `Cmod` equal to `Cmes` by trial and error, but we can be more systematic,

```{r}
age = seq(0,20e3,by=10)
Cmod = rep(NA,length(age))
for (i in 1:length(age)){
  Cmod[i] = solv_conc_eul(0,0,age[i],0,prm[,"Be10"],S*Ss,Lambda) # compute concentration
}
imin = which.min(abs(Cmod-Cmes))
res = age[imin]
plot(age,Cmod-Cmes,type="l",main=paste(res,"a BP"),col="coral",lwd=2)
abline(h=0)
abline(v=res)
```

> **TODO** Compare with the age from table 1 from @protin2019climatic. Comment about the possible causes for the differences?

We can be even more efficient by using an simple optimization approach to solve $C_{mod}(t) = C_{mes}$, using the built in `optimize` function. We just define a function to be optimized (search of minimum) and launch a search over an age range. We are looking for the minimum value of $|C_{mod}-C_{mes}|$

```{r}
fun_opt <-function(t,Cmes,prm,S,Lambda){
  Cmod = solv_conc_eul(0,0,t,0,prm[,"Be10"],S,Lambda)
  return(abs(Cmod-Cmes))
}
res = optimize(fun_opt,c(0,50e3),Cmes,prm,S*Ss,Lambda)
print(res)
```

## Time varying production rates
We are now going back to the time varying-scaling parameters

```{r}
data = data.frame(t1=seq(0,50e3,length.out=2000)) # we build a dataframe to store results
data$t2 = max(data$t1) - data$t1 # time BP
data$vdm = get_vdm(data$t2,model="musch")
data$rc = vdm2rc(data$vdm,latitude,model="elsasser54")
data$lm = scaling_lm(P,data$rc)
plot(data$t2,data$lm,type="l",xlab="Age BP (a)",ylab="Scaling factor")
abline(h=S$Nneutrons,col="red")
abline(h=mean(data$lm),lty=2)
```

In this case we can not use the same equations for the evolution of TCN  concentration as above, because we need to account for $P(t)$. We are going to use,
$$C_{mod}(t) = \int_0^t P(t)e^{-\lambda t}dt$$
Note : it is usually considered that production by muons is much less affected than that by neutrons by the magnetic field variations.

Below we first compare how the concentration evolves through time with the two approaches. 

```{r}
library(pracma) # useful library containing the cumtrapz function for trapezoidal integration
data$C1 = solv_conc_eul(0,0,data$t1,0,prm,S*Ss,Lambda) 
data$C2 = cumtrapz(data$t1,(prm[1]*data$lm*Ss+(prm[2,"Be10"] + prm[3,"Be10"])*Ss*S$Nmuons)*(exp(-prm["lambda",'Be10']*data$t1)))[,1] 
plot(data$t1,data$C1,type="l",xlab="time (a)",ylab="Concentration (at/g)",col="cyan4")
lines(data$t1,data$C2,col="darkgoldenrod2")
legend("topleft",c("st","lm"),lty=1,col=c("cyan4", "darkgoldenrod2"))
```

We can solve solve $|C_{mod}-C_{mes}|$ for $t$ with the same approach as the `st` case,

```{r}
compute_C<-function(t1,prm,P,lat,Lambda,S){
  data = data.frame(t1=seq(0,t1,length.out=1000)) # we build a dataframe to store results
  data$t2 = max(data$t1) - data$t1 # time BP
  data$vdm = get_vdm(data$t2,model="musch")
  data$rc = vdm2rc(data$vdm,lat,model="elsasser54")
  data$lm = scaling_lm(P,data$rc)
  Cmod = trapz(data$t1,(prm[1,"Be10"]*data$lm+(prm[2,"Be10"] + prm[3,"Be10"])*S$Nmuons)*(exp(-prm["lambda",'Be10']*data$t2)))
  return(Cmod)
}

fun_opt2 <-function(t1,Cmes,prm,P,lat,Lambda,S){
  Cmod =  compute_C(t1,prm,P,lat,Lambda,S)
  return(abs(Cmod-Cmes))
}

  P = atm_pressure(alt=altitude,lat=latitude,lon=longitude,model="era40") # compute atmospheric pressure at site
  S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
  res = optimize(fun_opt2,c(0,50e3),Cmes,prm,P,latitude,Lambda,S*Ss)
print(res)
```


Caveat we did those calculation using a number of simplification and numerical shortcuts ....

## Dealing with uncertainties