# Introduction

## Terrestrial Cosmogenic Nuclides and Earth Surface Dynamics

## Scope and objectives 

The objective of these activities is to allow users to explore a number of ideas and concepts related to the use of TCNs for the study of surface processes, in particular for exposure and denudation issues.

The book is built from small sequences of numerical code allowing to produce the presented figures. 
These sequences are organized to be easily reusable and allow to quickly test the influence of different parameters on the calculations.

## Working with this web site

This html site was generated from  [R Markdown](http://rmarkdown.rstudio.com) files using the [bookdown](https://bookdown.org) package.
You can copy/paste the various lines of code into your own R script and run it in any R session.

These activities do not require any specific prior knowledge of R programming language.
The idea is for you to simply copy and paste the code in a script, run it, and change various parameters to observe and investigate the associated response.

In addition to the scripting-oriented activities below you can also experiment visually with a few interactive [Shiny](https://shiny.rstudio.com) apps :

<http://shinyproxy.osupytheas.fr>

Both are built on the dedicated `TCNtools` package :

<https://vincentgodard.github.io/TCNtools>

This package contain various functions to assist the interpretation of TCN concentration at the Earth Surface.

The first thing you will have to do, before calling any of the functions, is always to load the `TCNtools` package.

```{r}
library("TCNtools")
```

But before that you need to have a R/RStudio working environment.

## Getting started with R/RStudio

Various options can be considered to run the R code bits presented in this book.

### Local installation

You can install [RStudio](https://www.rstudio.com) directly on your system and then the [TCNTools](https://github.com/VincentGodard/TCNtools) package using the instruction provided on the package webpage (requires `devtools` package to install from Github).

```{r eval=FALSE, include=TRUE}
install.packages("devtools")
devtools::install_git(url = "https://github.com/VincentGodard/TCNtools")
```

- **Pros**  
     - Run locally, not dependent on a distant server with possible usage limits
- **Cons** : 
     - Requires to install R/RStudio, which might not be interesting if you do not use them on a regular basis
     - `devtools` install can sometimes take some work, in particular on Windows systems  

### Binder

Alternatively you can launch a [binder](https://mybinder.org/) session and get a `RStudio` session running into your browser with the `TCNtools` package already installed, by clicking on the icon ![](images/binder.svg) on the package Github repository (top of Readme.md):

<https://github.com/VincentGodard/TCNtools>

- **Pros**  
     - Run into your browser
     - Nothing to install, just one click
- **Cons** : 
     - Temporary session, you loose everything when closing the browser (but it is possible to download your files) 
     - Limit on the number of instances that can run simultaneously
     

<!-- You can launch the binder by clicking on the icon ![](images/binder.svg) on the package website (right column). -->

### RStudio Cloud

Another option is to create a free account on [https://rstudio.cloud](https://rstudio.cloud).
Then start a project and install the `TCNTools` package.

```{r eval=FALSE, include=TRUE}
install.packages("devtools")
devtools::install_git(url = "https://github.com/VincentGodard/TCNtools")
```

- **Pros**  
     - Run in your browser
     - Simple installation 
     - Permanent account, allowing to keep projects and files from one session to the other
- **Cons** : 
     - A 25 hours per month limit with the free base plan. Remember to close you session when you are not working


## Usage


Various boxes will be used throughout the book.

:::: {.codebox .codeimage data-latex="code"}
This a summary code box. 
In the following chapters the processes will be explained step by step, with the various actions distributed between different successive code blocks (also named chunks) for better understanding.

For convenience, and avoid you to have to copy and paste these various blocks one by one, sometimes everything will be combined in one block at the end of the section, which can unfold, copy and then paste into the R console or a script.
Below you can get all the necessary code to produce a figure or result.
```{r eval=FALSE, class.source='fold-hide', include=TRUE}
a = 1
b = 3
c = (a+b)^2
```
::::


:::: {.actionbox .actionimage data-latex="action"}
The purpose of this book is to allow you to experiment and explore, not just copy and paste some code.
This kind of box will suggest you some action on the code, typically changing the value of parameters to observe the evolution of the result.
::::

:::: {.notebox .noteimage data-latex="note"}
This box will be used to highlight or stress something important.
::::

:::: {.questionbox .questionimage data-latex="question"}
This box will be used to formulate some questions for you to think about.
::::



