# Application to exposure dating

```{r ed0, include=FALSE}
library("TCNtools")
data(Lambda)
data(prm)
rho = 2.7
#
```

## Back to the evolution of concentration

We first consider the evolution of concentration with time $t$.
The computation will be carried out at the surface ($z=0$), but this could be done at any arbitrary depth.
We also consider that $\varepsilon = 0$ and that there is no inheritance.
The starting equation above becomes, 
\begin{equation}
C(t)=\sum_i \frac{P_i}{\lambda}(1-e^{-\lambda t}) (\#eq:pureexpo)
\end{equation}

For the sake of the example we are going to compute that by hand, and compare with the results obtained with the `solv_conc_eul` function,

Again we define the usual parameters 

```{r ed1}
# set up
altitude = 1000 # elevation in m
latitude = 45 # latitude in degrees
nuc = "Be10"
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

To improve readability we compute scaled production rates and give them explicit names.

```{r ed2}
Pspal = prm[1,nuc]*S$Nneutrons # scaled spallation production rate in at/g/y (st scaling)
Pstop = prm[2,nuc]*S$Nmuons # scaled stopped muons  production rate in at/g/y
Pfast = prm[3,nuc]*S$Nmuons # scaled fast muons  production rate in at/g/y
lambda = prm[4,nuc] # radioactive decay (1/y)
```

And then apply equation \@ref(eq:pureexpo) and compare the results with the output of function `solv_conc_eul`.

```{r ed3}
t =  seq(0,200e3,by=100) # a vector containing time from 0 to 100 ka by 100 a steps
C = (Pspal + Pstop + Pfast) / lambda * (1-exp(-lambda*t))
plot(t/1000,C,type="l",col="coral",lwd=3,ylab="Concentration (at/g)",xlab="Time (ka)")
grid()
# check if this is ok
C2 = solv_conc_eul(0,0,t,0,prm[,nuc],S,Lambda) # compute concentration for pure exposure
lines(t/1000,C2,lty=2)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('ed0','ed1','ed2','ed3')}
```
::::

One thing to note in this graph is that time $t$ is expressed as *time elapsed since start of exposure*, which we can not directly translate into the age BP, when dealing with radioactive decay and time variations in production rate. 

:::: {.actionbox .actionimage data-latex="action"}
Change the maximum time `t` to check that everything is correct.
::::


## Interpretation in terms of exposure ages

We are now going to make some simple calculations to interpret concentrations in terms of exposure age. 
The objective is solving for $t$ the equation $C_{mod}(t) = C_{mes}$. 
The left hand side is our modelled concentration, obtained for example with equation \@ref(eq:conceuler) (`solv_conc_eul`), where we need to keep in mind all the hypotheses we make ($z=0$, $\varepsilon=0$, $C_0=0$). 

We are going to use an example from @protin2019climatic, using sample *ARG-16-9*, which was collected from a lateral moraine of the Argentière glacier (Northern French Alps).

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(leaflet)
df = data.frame(X =  6.96065,Y = 45.96670,label="ARG-16-9")
  
m <- leaflet(df) %>%
  addTiles() %>%
  addProviderTiles("OpenStreetMap",group = "OpenStreetMap") %>%
  addProviderTiles("Esri.WorldImagery",group = "Esri.WorldImagery") %>%
  addLayersControl(baseGroups = c("OpenStreetMap","Esri.WorldImagery"),position = "topleft") %>%
  addLabelOnlyMarkers(data = df,
                      lng = ~X, lat = ~Y, label = ~label, 
                      labelOptions = labelOptions(noHide = TRUE, direction = 'auto', textOnly = FALSE)) 
m
```

Here are the characteristics of this sample (see table 1 from this paper).

```{r ed4}
rho = 2.7
altitude = 2252 # elevation in m
latitude = 45.97 # latitude in degrees
longitude = 6.96 # longitude in degrees
Ss = 0.939*exp(-2.6*rho/Lambda[1]) # factor accounting for topographic shielding and sample thickness
Cmes = 26.8e4 # measured concentration (at/g)
Cmes_e = 1.3e4
#
P = atm_pressure(alt=altitude,lat=latitude,lon=longitude,model="era40") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

Now, for example we can make some guess about the exposure age and see what is the difference with the measured concentration.

```{r ed5}
age = 10e3 # a 
Cmod = solv_conc_eul(0,0,age,0,prm[,"Be10"],S*Ss,Lambda) # compute concentration
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to compute `Cmod`
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('ed0','ed4','ed5')}
```
::::

:::: {.actionbox .actionimage data-latex="action"}
Change the value of age and try to get close to measured $^{10}$Be concentration.
::::

We could try make `Cmod` equal to `Cmes` by trial and error, but we can be more systematic. 
We go through automatically over a whole range of `age` values, and compute the corresponding `Cmod` values.

```{r ed6}
age = seq(0,20e3,by=10) # we define a vector for age, which set the different explored values
Cmod = rep(NA,length(age)) # The vector of same length to store the results
for (i in 1:length(age)){
  Cmod[i] = solv_conc_eul(0,0,age[i],0,prm[,"Be10"],S*Ss,Lambda) # compute concentration
}

```

We then plot the results to identify the minimum value of $|C_{mod} - C_{mes}|$

```{r ed7}
imin = which.min(abs(Cmod-Cmes)) # what is the index of the minimum difference
res = age[imin] # the corresponding age
plot(age,Cmod-Cmes,type="l",main=paste(res,"a BP"),col="coral",lwd=2)
abline(h=0)
abline(v=res)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('ed0','ed4','ed5','ed6','ed7')}
```
::::

:::: {.questionbox .questionimage data-latex="action"}
Compare with the age reported in table 1 from @protin2019climatic. Comment about the possible causes for the differences?
::::


We can be even more efficient by using an simple optimization approach to solve $C_{mod}(t) = C_{mes}$, using the built in `optimize` function. 
We just define a function to be optimized (search of minimum) and launch a search over an age range. 
We are looking for the minimum value of $|C_{mod}(t)-C_{mes}|$. 
Here the function we are trying to optimize.

```{r ed8}
fun_opt <-function(t,Cmes,prm,S,Lambda){
  Cmod = solv_conc_eul(0,0,t,0,prm[,"Be10"],S,Lambda)
  return(abs(Cmod-Cmes))
}
```

We then launch the search of the 0-50 ka interval.

```{r ed9}
res = optimize(fun_opt,c(0,50e3),Cmes,prm,S*Ss,Lambda)
print(res)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to obtain this result
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('ed0','ed4','ed8','ed9')}
```
::::

:::: {.questionbox .questionimage data-latex="action"}
This is a very coarse calculation, but searching the  solution for $C_{mod}(t)=C_{mes}$ is the essence of what online calculator do.
What kind of improvement could you think of concerning this calculation?  
::::


## Time varying production rates

We are now going back to the time varying-scaling schemes, such as the one we explored in the previous chapter.

```{r ed10}
data = data.frame(t1=seq(0,20e3,length.out=2000)) # we build a dataframe to store results
data$t2 = max(data$t1) - data$t1 # time BP
data$vdm = get_vdm(data$t2,model="musch")
data$rc = vdm2rc(data$vdm,latitude,model="elsasser54")
data$lm = scaling_lm(P,data$rc)
plot(data$t2,data$lm,type="l",xlab="Age BP (a)",ylab="Scaling factor")
abline(h=S$Nneutrons,col="red")
abline(h=mean(data$lm),lty=2)
```

In this case we can not use the same equations for the evolution of TCN  concentration as above, because we need to account for $P(t)$. 
We are going to use,
$$C_{mod}(t) = e^{-\lambda t} \int_0^t P(t')e^{\lambda t'}dt'$$

Note : it is usually considered that production by muons is much less affected than that by neutrons by the magnetic field variations.

Below we first compare how the concentration evolves through time with the two approaches. 

```{r ed11}
library(pracma) # useful library containing the cumtrapz function for trapezoidal integration
#
Pspal = prm[1,"Be10"]*S$Nneutrons*Ss # scaled spallation production rate in at/g/y (st scaling)
Pstop = prm[2,"Be10"]*S$Nmuons*Ss # scaled stopped muons  production rate in at/g/y
Pfast = prm[3,"Be10"]*S$Nmuons*Ss # scaled fast muons  production rate in at/g/y
lambda = prm[4,"Be10"] # radioactive decay (1/y)
data$Prod_st = Pspal + Pstop + Pfast
data$C1 = solv_conc_eul(0,0,data$t1,0,prm,S*Ss,Lambda) 
#
data$Prod_lm = prm[1,"Be10"]*data$lm*Ss+(prm[2,"Be10"] + prm[3,"Be10"])*Ss*S$Nmuons
data$C2 = exp(-lambda*data$t1)*cumtrapz(data$t1,data$Prod_lm*exp(lambda*data$t1))
#
plot(data$t1,data$C1,type="l",xlab="Time since start of exposure (a)",ylab="Concentration (at/g)",col="cyan4")
lines(data$t1,data$C2,col="darkgoldenrod2")
legend("topleft",c("st","lm"),lty=1,col=c("cyan4", "darkgoldenrod2"))
```

:::: {.questionbox .questionimage data-latex="question"}
Comment on the differences.
::::

We can solve solve $C_{mod}(t)=C_{mes}$ for $t$ with the same approach as the `st` case,

```{r ed12}
library(pracma) # useful library containing the cumtrapz function for trapezoidal integration
#
compute_C<-function(t1,prm,P,lat,Lambda,S,Ss){
  data = data.frame(t1=seq(0,t1,length.out=1000)) # we build a dataframe to store results
  data$t2 = max(data$t1) - data$t1 # time BP
  data$vdm = get_vdm(data$t2,model="musch")
  data$rc = vdm2rc(data$vdm,lat,model="elsasser54")
  data$lm = scaling_lm(P,data$rc)*Ss
  Cmod = exp(-prm["lambda",'Be10']*data$t1[nrow(data)])*trapz(data$t1,(prm[1,"Be10"]*data$lm+(prm[2,"Be10"] + prm[3,"Be10"])*S$Nmuons)*(exp(prm["lambda",'Be10']*data$t1)))
  return(Cmod)
}
#
fun_opt2 <-function(t1,Cmes,prm,P,lat,Lambda,S,Ss){
  Cmod =  compute_C(t1,prm,P,lat,Lambda,S,Ss)
  return(abs(Cmod-Cmes))
}
#
  P = atm_pressure(alt=altitude,lat=latitude,lon=longitude,model="era40") # compute atmospheric pressure at site
  S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
  res = optimize(fun_opt2,c(0,50e3),Cmes,prm,P,latitude,Lambda,S,Ss)
print(res)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to obtain this result
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('ed0','ed4','ed12')}
```
::::


Caveat we did those calculation using a number of simplification and numerical shortcuts ....

## Dealing with uncertainties