# Two nuclides systems

```{r include=FALSE}
library("TCNtools")
```

We first define the usual variables and parameters.

```{r }
data(Lambda) # we load a vector containing the attenuation length into the environment
data(prm) # we load a matrix containing the production/decay parameters into the environment
rho = 2.7 # we also define the density (g/cm3)
altitude = 1000 # elevation in m
latitude = 45 # latitude in degrees
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

Here first we consider that all samples are close and have similar positions and elevation, so similar scaling parameters.
We will see later how to deal with situations where samples have different positions and hence different scaling parameters.

In order to avoid messing up things later we define here once and for all which is nuclide 1 and which is nuclide 2, according to their respective $\tau_{1/2}$ :

```{r}
N1 = "Be10" # longer half-life
N2 = "Al26" # shorter half-life
```


## Evolution of ratios

We first start by looking into the evolution of the isotopic ratio between the nuclides, once production has stopped, which is one of the key phenomenon used in burial dating applications.

We will look at the evolution through time (typically over several Ma for the 26/10 system), starting from concentrations defined by a denudation steady-state.

```{r}
data = data.frame(t = seq(0,7e6,by=2000))
ero = 50 * rho * 100 / 1e6 # in g/cm2/a
```

We compute the initial steady-state concentrations.

```{r}
C1_0 = solv_conc_eul(0,ero,Inf,0,prm[,N1],S,Lambda)
C2_0 = solv_conc_eul(0,ero,Inf,0,prm[,N2],S,Lambda)
```

Now we let them decay through time starting from the initial values, with no production ($z=+\infty$).
We plot the evolution of the ratio through time. 

```{r}
data$C1 = solv_conc_eul(Inf,0,data$t,C1_0,prm[,N1],S,Lambda)
data$C2 = solv_conc_eul(Inf,0,data$t,C2_0,prm[,N2],S,Lambda)
plot(data$t,data$C2/data$C1,type="l",lwd=2,col="darkorange",xlab="Time (a)",ylab=paste(N2,"/",N1))
```





## Steady-state denudation and constant exposure curves

Two-nuclides plots are usually built around two curves representing the predictions, in terms of concentrations, for end-member simplified situations :

- **steady-state denudation**, where the concentrations of the nuclides of interest are computed for various values of surface denudation rates
- **constant exposure**, where the concentrations of the nuclides of interest are computed for various duration of surface exposition, with no denudation

We will use the Eulerian point of view (function `solv_conc_eul`) to compute the concentrations, which is done by function `tnp_curves`.

We compute the data for these two curves.

```{r}
tmp = tnp_curves(prm[,N1],prm[,N2],Lambda,S,rho)
ss_ero = tmp[[1]]
cst_exp = tmp[[2]]
```

The default parameters for the ranges of denudation rates and exposure ages can be adjusted if needed.

Now that we can plot everything in a two-nuclides graph.
While not mandatory, it is usually recommended to organize the plot this way :

- **X-axis** : longer half-life nuclide (our $N_1$)
- **Y-axis** : ratio of shorter to longer half-life nuclides (our $N_2/N_1$)

```{r test}
plot(NA,xlim=range(cst_exp$C1,ss_ero$C1),ylim=range(cst_exp$C2/cst_exp$C1,ss_ero$C2/ss_ero$C1),log="x",
     xlab=paste(N1,"(at/g)"),ylab=paste(N2,"/",N1))
lines(cst_exp$C1,cst_exp$C2/cst_exp$C1,lty=2,col="khaki4") # constant exposure, dashed line
lines(ss_ero$C1,ss_ero$C2/ss_ero$C1,col="khaki4") # steady-state erosion, solid line
lines(data$C1,data$C2/data$C1,lwd=2,col="darkorange")
```

We have also plotted the burial trajectory we generated above.
Starting from a position on the steady-state denudation curve, we progressively lower the ratios.

> **TODO** Change the denudation rate used to compute the initial steayd-state concentrations

## Interactive shiny app

You can also explore dynamically 2 nuclides plots using this embedded application, which use the same type of code. 

```{r echo=FALSE}
knitr::include_url("https://shinyproxy.osupytheas.fr/2_nuclides/",height = "1200px")
```


