# Exploring TCN build up at the surface

```{r bu0, include=FALSE}
library("TCNtools")
data(Lambda)
data(prm)
rho = 2.7
#
```


We are going to consider simple computations of concentration under various conditions in terms of erosion, depth or age.
This will be done using an *Eulerian point of view*, which is the most straightforward and fastest way to perform such computation.
In this case the quantity of interest (concentration) is computed at fixed depths below the surface, while the exhumed material is moving through this reference frame during its trajectory toward the surface.
More details on the differences between Eulerian and Lagrangian approaches, and their applications to complex exposition/denudation histories, will be presented later.

Note that interpreting measured concentrations in terms of end-member situations of *pure exposure* or *steady-state denudation* is often done with online calculators (@balco2008complete,@marrero2016cosmogenic,@martin2017crep).
Those calculators allow to perform very accurate computations of age or denudation rate, but one should always be careful about the underlying hypotheses (no erosion, steady state achieved, etc ...) when interpreting measured concentrations and always think about how TCN are accumulating in the sampled material.

The goal of this activity is to explore this behavior, using simple approaches.

## Background

The relevant general equation is the following,

\begin{equation}
C=C_0e^{-\lambda t} + \sum_i \frac{P_i}{\frac{\rho \varepsilon}{\Lambda_i}+\lambda}e^{\frac{-\rho z}{\Lambda_i}}(1-e^{-(\frac{\rho \varepsilon}{\Lambda_i}+\lambda)t}) (\#eq:conceuler)
\end{equation}

with the following variables and parameters,

-   $C$ the concentration (as a function of time $t$ and depth $z$)
-   $C_0$ the inherited concentration
-   $\lambda$ the decay constant for the considered nuclide
-   $P_i$ the scaled surface production rate for the nuclide of interest and the $i$-th production pathway (spallation, stopped muons, fast muons)
-   $\rho$ the density of the medium
-   $\Lambda_i$ the attenuation length for the particules of the $i$-th production pathway
-   $\varepsilon$ surface denudation

In order to stick with usual conventions in the following time $t$ will be measured in years (a), the unit of length will be cm and the depths ($z$) will be expressed in g/cm$^2$ (i.e. actual depth $\times \rho$).

Note two keys limitations of this representation :

-   it does not allow to account for time variations of production rates (at least in its most straightforward implementation), so we will mostly using the *st* scaling
-   it assumes exponential evolution of production with depth, which is clearly not the case for low energy neutrons (figure 2b from @gosse2001terrestrial) and is questionable in some situations for muons (@balco2017production)

## Set up of the calculations

We should introduce some of the basic parameters we are going to use for the computation. 
For easy reference a set of data for parameters of interest is included in the `TCNtools` package.

-   a vector (`Lambda`) with the attenuation lengths for different particles (in g/cm$^2$)
    -   neutrons for spallation reactions $\Lambda_{spal}$
    -   stopping muons $\Lambda_{stop}$
    -   fast muons $\Lambda_{fast}$\
-   a vector (`prm`) with the SLHL production rates (in at/g/a), in this case for the *st* scaling scheme (@stone2000air), and decay constant $\lambda$ (in 1/a) for the nuclides of interest. 
Note that this will need to be modified when using other scaling schemes.

Note that these often used SLHL values are defined for convenience, most calculators for exposure age work directly with the production rate value at calibration sites, and that they are always relative to the scaling scheme used (@borchers2016geological).

We can first load the attenuation length data (g/cm$^2$).
Documentation of this dataset is accessible by typing `?Lambda` in the R console.

```{r bu1}
data(Lambda) # we load a vector containing the attenuation length into the environment
print(Lambda)
rho = 2.7 # we also define the density (g/cm3)
```

Some production and decay parameters can also be loaded.
Documentation of this dataset is accessible with `?prm`.

```{r bu2}
data(prm) # we load a matrix containing the production/decay parameters into the environment
print(prm)
```

We also need to define the properties of our site of interest and compute the relevant scaling parameters. 
As we already saw previously, this can easily be done with,

```{r bu3}
altitude = 1000 # elevation in m
latitude = 45 # latitude in degrees
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

## Evolution of concentration with time

To get a general overview of the behavior we are  going to use directly the `solv_conc_eul` function, which allows to easily deal with various scenarios and configurations, using equation \@ref(eq:conceuler). 
In the following chapters we will go back to the key equations to get a better sense of the importance of various parameters.
As always the documentation of the function, including its various arguments, can be obtained by typing `?solv_conc_eul` in the R console.

We start by defining the various 

```{r bu4}
nuc = "Be10" # "Al26", "C14"
t =  seq(0,200e3,length.out=1000) # a vector containing time from 0 to 100 ka by 100 a steps
z = 0 * rho # depth at which we are going to perform the calculation (cm converted to g/cm2)
C0 = 0 # inherited concentration (at/g)
ero = 0 * (100/1e6*rho) # denudation rate expressed in m/Ma and converted in g/cm2/a
```

Now we can compute the concentration (at various times `t`), according to equation \@ref(eq:conceuler).

```{r bu5}
C = solv_conc_eul(z,ero,t,C0,prm[,nuc],S,Lambda) # compute concentration
```

Then we can plot the evolution of concentration with time.

```{r bu6}
plot(t/1000,C,type="l",col="cornflowerblue",lwd=3,ylab="Concentration (at/g)",xlab="Time (ka)")
grid()
# see explanation for these  lines below
Prod = c(prm[1,nuc]*S$Nneutrons,prm[2,nuc]*S$Nmuons,prm[3,nuc]*S$Nmuons) # scaled production vector (defined for the sake of clarity of the expressions below)
lambda = prm[4,nuc] # radiactive decay
abline(0,sum(Prod)*1000,lty=2) # note that time is in ka on the plot
abline(h=sum(Prod/((ero/Lambda)+lambda)),lty=2)
```

:::: {.codebox .codeimage data-latex="caution"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('bu0','bu3','bu4','bu5','bu6')}
```
::::

Starting with zero erosion ($\varepsilon=0$), corresponding to the pure exposition of a surface, we see the progressive build-up of concentration though time and the establishment of a balance between gains (production) and losses (denudation and decay) leading to the concentration plateau at steady state.

Note that two dashed lines are added to the graph,

- The first one corresponds to the production slope $\sum_i P_i$, how much nuclide you produce and how it would accumulate if you had no radioactive decay and no denudation.
- The second one (horizontal, not visible with the initial parameters) is the maximum value of concentration when steady state is achieved :
$$ C_{max}=\sum_i \frac{P_i}{\frac{\rho \varepsilon}{\Lambda_i}+\lambda} $$

:::: {.actionbox .actionimage data-latex="action"}
-   Change the maximum of the `t` vector until you see the influence of radioactive decay and the plateau.
-   Add some inheritance ($C_0$)
-   Test the evolution with other nuclides such as $^{26}$Al and $^{14}$C (set `nuc` to `Al26` or `C14`)
-   Modify the `ero`  parameter above (always keeping it is in g/cm$^2$/a), to see its influence on time needed to reach steady state and the final concentration
::::


## Two end-member situations

Now we are going to build a summary plot showing the influence of both exposure and denudation.


```{r bu7}
nuc = "Be10" # choice of nuclide
t =  10^seq(log10(1),log10(10e6),length.out=1000) # time vector, log-spaced!
# calculation of the evolution of concentration for denudation = 0
C = solv_conc_eul(0,0,t,0,prm[,nuc],S,Lambda) # compute concentration for pure exposure
plot(t,C,type="l",col="cornflowerblue",lwd=3,ylab="Concentration (at/g)",xlab="Time (a)",log="xy")
grid()
text(max(t),max(C),0,cex=0.5,adj=0) # label the curve
# now we make the same computation for other denudation rates (using a loop)
ero = c(1,10,100,1000) # erosion vector in m/Ma
for (i in 1:length(ero)){
  e = ero[i] * (100/1e6*rho) # convert denudation in g/cm2/a
  C = solv_conc_eul(0,e,t,0,prm[,nuc],S,Lambda) # compute concentration for pure exposure
  lines(t,C,col="cornflowerblue",lwd=3)
  text(max(t),max(C),ero[i],cex=0.5,adj=0) # label the curve
}
```

:::: {.codebox .codeimage data-latex="caution"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('bu0','bu3','bu7')}
```
::::


Note that this a log-log plot. 
It is probably one of the most important figure to keep in mind when analyzing TCN concentrations. 
It clearly shows the existence of two end-member situations when interpreting these concentrations, in terms of exposure age or denudation rates, and the transition between the two. 

:::: {.questionbox .questionimage data-latex="action"}
Think about a bit about the following points

- What are the key hypotheses made when interpreting a TCN concentration in terms of
    - exposure age
    - surface denudation
- Can you think of geological/geomorphological situations where these hypotheses are invalid?
- Why are the plateau concentrations so different?
- How is the time to reach the plateau changing and why?
::::



## Comparing the Eulerian and Lagrangian descriptions

This section presents a simple comparison of the Eulerian and Lagrangian approaches for the computation of cosmogenic nuclides concentration evolution in complex denudation and exposure scenarios.
A nice presentation of this distinction is provided in @knudsen2019timeintegrating.

### Definition of the denudation scenario

We will start from a steady state situation under constant denudation, and then apply a change in the denudation rate for a given period of time.

```{r bu11}
tmax = 2e6 # duration in a 
ero = 1/1.e6*100*rho # denudation rate conversion en m/Ma -> g/cm2/a
fact = 5 # change factor
```

Here the duration of the simulation will be `r tmax/1e6` Ma, the initial denudation rate will be `r ero*1e6/100/rho` m/Ma, which will changed by a factor `r fact` at the beginning of the simulation.

### Computation of concentrations with the Eulerian reference frame

We compute the evolution of concentrations at the surface ($z=0$) using an Eulerian point of view.
The calculation is very straightforward with to the `solv_conc_eul` function, note the use of `in_ero` parameter, to specify a starting denudation rate.

```{r bu12}
df_e = data.frame(t = seq(0,tmax,length.out = 10000))
df_e$C10 = solv_conc_eul(0,ero*fact,df_e$t,0,prm[,"Be10"],S,Lambda,in_ero=ero)
```

### Computation of concentrations with the  Lagrangian reference frame

Now we do the same calculation using a Lagrangian point of view, i.e. with a reference frame attached to a rock particle during its journey toward the surface as a response to denudation. 
It is slightly more complicated than in the Eulerian case and is done using the `solv_conc_lag` function.

We first initiate a dataframe (`df_l`) and compute the evolution in depth `z` of the particle through time when it is exhumed from its initial depth (here `r ero*fact*tmax/rho/100` m) to the surface at a rate `r ero*fact/rho/100*1e6` m/Ma.

```{r bu13}
df_l = data.frame(t=seq(0,tmax,length.out = 10000)) # data frame to store results
df_l$z = ero*fact*df_l$t # cumulative erosion trough time (g/cm2)
df_l$z = max(df_l$z) - df_l$z # convert into depth
```

We compute the steady state concentration at starting depth, using the Eulerian solver.

```{r bu14}
C10_0 = solv_conc_eul(max(df_l$z),ero,Inf,0,prm[,'Be10'],S,Lambda) # starting 10Be concentration
```


We then define another column in the data frame with the evolution of scaling parameters through time. 
It is trivial here in the case of the time-independent **st** scaling, but it will be highly valuable when dealing with time-dependent scalings, where the Eulerian approach is not applicable.
 
```{r bu15}
df_l$Ssp  = rep(as.numeric(S[1]),nrow(df_l)) # scaling spallation
df_l$Smu  = rep(as.numeric(S[2]),nrow(df_l)) # scaling muons
```

Then we calculate the un-scaled production rates at depths  of interest. 
Here again, as we use simple models of exponential decrease for production with depth, the interest of using such Lagrangian approach is not obvious. 
But such way of computing concentration will allow to deal with non-exponential production profiles, which might necessary in some situation for  muons or low-energy neutrons. 


```{r bu16}
# 10Be (not scaled)
df_l$Psp10 = prm["Pspal",'Be10']*exp(-1*df_l$z/Lambda["Lspal"]) # spallation
df_l$Pmu10 = prm["Pstop",'Be10']*exp(-1*df_l$z/Lambda["Lstop"]) + prm["Pfast",'Be10']*exp(-1*df_l$z/Lambda["Lfast"]) # muons
```

We can now use the `solv_conc_lag` function to compute the evolution of concentration through time and depth.
```{r bu17}
df_l$C10 = solv_conc_lag(df_l$t,df_l$z,C10_0,df_l$Psp10,df_l$Pmu10,prm["lambda",'Be10'],cbind(df_l$Ssp,df_l$Smu),final=FALSE)
```

### Comparison of results

We can now plot the results of the calculations and compare the two approaches.

```{r bu18}
col_e = "chartreuse"
col_l = "chocolate1"
  
plot(df_e$t/1e6,df_e$C10/1e6,type="l",col=col_e,lwd=2, xlim=range(df_e$t/1e6),ylim=range(df_e$C10/1e6,df_l$C10/1e6),
     xlab="Time (Ma)",ylab="10Be concentration (x10e6 at/g)")
grid()
lines(df_l$t/1e6,df_l$C10/1e6,col=col_l,lwd=2)
legend("topright",c("Eulerian","Lagrangian"),lwd=2,col=c(col_e,col_l))
```

:::: {.codebox .codeimage data-latex="caution"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('bu0','bu3','bu11','bu12','bu13','bu14','bu15','bu16','bu17','bu18')}
```
::::

The Lagrangian evolution starts with a low concentration corresponding to the burial at depth of the particle at the onset of the evolution, while at the same time ($t=0$) we are looking at the surface with the Eulerian point of view, and then obviously observe much higher concentrations.
The two descriptions converge toward the same concentration at the end of the evolution (i.e. when the lagrangian particle reaches the surface).

:::: {.questionbox .questionimage data-latex="question"}
Try to think about various contexts and applications where the use of the Eulerian or Lagrangian appraoch could be preferred, and why.
::::

## Interactive shiny app

You can also explore dynamically the TCN build up using this embedded application, which use the same type of code. 

```{r echo=FALSE}
knitr::include_url("https://shinyproxy.osupytheas.fr/simple_accumulation/",height = "1200px")
```


