# Two nuclides systems

```{r include=FALSE}
library("TCNtools")
```

We first define the usual variables and parameters

```{r }
data(Lambda) # we load a vector containing the attenuation length into the environment
data(prm) # we load a matrix containing the production/decay parameters into the environment
rho = 2.7 # we also define the density (g/cm3)
altitude = 1000 # elevation in m
latitude = 45 # latitude in degrees
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

Here first we consider that all samples are close and have similar positions, so similar scaling parameters.
We will see later how to deal with situations where samples have different positions and hence different scaling parameters.

## Steady-state denudation and constant exposure curves

Two-nuclides plots are usually built around two curves representing the predictions, in terms of concentrations, for end-member simplified situations :

- **steady-state denudation**, where the concentrations of the nuclides of interest are computed for various values of surface denudation rates
- **constant exposure**, where the concentrations of the nuclides of interest are computed for various duration of surface exposition, with no denudation

We will use the Eulerian point of view (function `solv_conc_eul`) to compute the concentrations, which is done by function `tnp_curves`.

In order to avoid messing up things later we define here once and for all which is nuclide 1 and which is nuclide 2, according to their respective $\tau_{1/2}$ :

```{r}
N1 = "Be10" # longer half-life
N2 = "Al26" # shorter half-life
```

We compute the data for these two curves.

```{r}
tmp = tnp_curves(prm[,N1],prm[,N2],Lambda,S,rho)
ss_ero = tmp[[1]]
cst_exp = tmp[[2]]
```

The default parameters for the ranges of denudation rates and exposure ages can be adjusted if needed.

Now that we can plot everything in a two-nuclides graph.
While not mandatory, it is usually recommended to organize the plot this way :

- **X-axis** : longer half-life nuclide (our $N_1$)
- **Y-axis** : ratio of shorter to longer half-life nuclides (our $N_2/N_1$)

```{r test}
plot(NA,xlim=range(cst_exp$C1,ss_ero$C1),ylim=range(cst_exp$C2/cst_exp$C1,ss_ero$C2/ss_ero$C1),log="x",
     xlab=paste(N1,"(at/g)"),ylab=paste(N2,"/",N1))
lines(cst_exp$C1,cst_exp$C2/cst_exp$C1,lty=2,col="khaki4") # constant exposure, dashed line
lines(ss_ero$C1,ss_ero$C2/ss_ero$C1,col="khaki4") # steady-state erosion, solid line
```

The extent of the plot can be modified to correspond more closely to the reported data.


