# Application to denudation rate measurements


```{r dm0, include=FALSE}
library("TCNtools")
```


## Concentration - denudation rate relationship

Now we are going to consider the evolution of concentration with denudation rate $\varepsilon$.
The computation will be carried out at the surface ($z=0$), but this could be done at any arbitrary depth.
We will consider that $t=+\infty$ and that we have reached the plateau concentration.
Equation \@ref(eq:conceuler) becomes,

\begin{equation}
C=\sum_i \frac{P_i}{\frac{\rho \varepsilon}{\Lambda_i}+\lambda} (\#eq:ssero)
\end{equation}

Which simplifies to $C\approx\frac{P_i \Lambda_i}{\rho \varepsilon}$ if we neglect radioactive decay.

We define the usual parameters.

```{r dm1}
altitude = 1000 # elevation in m
latitude = 45 # latitude in degrees
rho = 2.7
data(prm)
data(Lambda)
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

Now compute the steady-state concentration for a range of denudation rates.

```{r dm2}
nuc = "Be10"
ero = 10^seq(log10(0.1),log10(1000),length.out = 100) * 100/1e6*rho # a log-spaced vector for denudation rate expressed in m/Ma and converted in g/cm2/a
age =  Inf # infinite age
z = 0 * rho # depth at which we are going to perform the calculation (cm converted to g/cm2)
C0 = 0 # inherited concentration
C = solv_conc_eul(z,ero,age,C0,prm[,nuc],S,Lambda) # compute concentration
```

We plot the result, and also consider consider the implications of neglecting the radioactive decay.

```{r dm3}
plot(ero/100*1e6/rho,C,col="lawngreen",log="xy",type="l",lwd=3,ylab="Concentration (at/g)",xlab="Denudation rate (m/Ma)")
# what happens if we neglect radioactive decay
Prod = c(prm[1,nuc]*S$Nneutrons,prm[2,nuc]*S$Nmuons,prm[3,nuc]*S$Nmuons) # scaled production vector (defined for the sake of clarity of the expressions)
lambda = prm[4,nuc] # radioactive decay
C2 = sum(Prod*Lambda)/ero
lines(ero/100*1e6/rho,C2,lty=2)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('dm0','dm1','dm2','dm3')}
```
::::


This figure (log-scales on both axes) highlights the strong inverse relationship, at steady-state, between denudation rate ($\varepsilon$) and concentration ($C$), which is the foundation of many geomorphological studies trying to establish landscape evolution rates.
Note the change in the relationship at very low denudation rates, which corresponds to the situation where the effects of radioactive decay become predominant.

:::: {.questionbox .questionimage data-latex="question"}
Over what range of denudation rates it is reasonable to neglect radioactive decay? What kind of geological context could it correspond to?
::::

A simple way to answer this question would be to compute the relative difference between the computed concentrations

```{r dm4}
error = abs(C-C2)/C*100
plot(ero/100*1e6/rho,error,log="xy",type="l",ylab="Relative error (%)",xlab="Denudation rate (m/Ma)")
```


## Integration time scale


## Catchment wide denudation rates
It can be noted that  when dealing with eroding bedrock surfaces the calculated denudation rate will be a very local estimate and that its extrapolation to longer wavelengths is only possible in the rather limited cases where the morphological properties of the relief are uniform over large distances (plateaus, morphological surfaces, etc ...). 
In reality, in complex environments, such as mountain ranges, where several processes contribute to denudation, a simple local estimate does not provide really useful information on the overall dynamics of the topography. 
An alternative approach is to use river sediments, which are a mixture of contributions from different parts of a watershed. 
The measurement of a cosmogenic nuclide concentration in this type of sample provides an estimate of the average denudation over the corresponding basin (@brown1995denudation,@granger1996spatially,@vonblanckenburg2005control).

Calculating such average denudation rate requires to obtain an average production rate for the catchment.
For low relief catchments the scaling parameters can be calculated using the average elevation of the catchment. But for catchments with several 1000s m in relief, the non linear relationship between the scaling parameters and elevation will make this approximation invalid.

We are going to explore these ideas using 3 basins from the study of @godard2012impact in the Marsyandi river basin (central Nepal)


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(leaflet)
bas0 <- st_read("data/gis/marsyandi.gpkg", quiet = TRUE)
bas <- st_transform(bas0, 4326)
centers <- st_centroid(bas)
xy = st_coordinates(centers)
m <- leaflet(bas) %>%
  addTiles() %>%
  addProviderTiles("OpenStreetMap",group = "OpenStreetMap") %>%
  addProviderTiles("Stamen.Terrain",group = "Stamen.Terrain") %>%
  addProviderTiles("Esri.WorldImagery",group = "Esri.WorldImagery") %>%
  addLayersControl(baseGroups = c("OpenStreetMap","Stamen.Terrain", "Esri.WorldImagery"),position = "topleft") %>%
  addPolygons() %>%
  addLabelOnlyMarkers(data = centers,
                    lng = xy[,1], lat = xy[,2], label = ~BASIN, 
                    labelOptions = labelOptions(noHide = TRUE, direction = 'auto', textOnly = FALSE)) 
m
```


Here are the maps for these 3 basins.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library("terra",verbose = F)
bas = vect("data/gis/marsyandi.gpkg")
layout(matrix(c(1,2,3), 1, 3, byrow = TRUE))
# plot ----
for (name in c("Khudi","Chudi","Marsyandi")){
dem = rast(paste0("data/gis/",name,".tif"))
slope <- terra::terrain(dem, "slope", unit="radians")
aspect <- terra::terrain(dem, "aspect", unit="radians")
hill <- terra::shade(slope, aspect, 40, 270)
plot(hill, col=grey(0:100/100), legend=FALSE,axes=F,main=name,mar=c(4, 2, 4, 4))
leg = terrain.colors(100,alpha=0.35)
plot(dem, col=leg, add=TRUE,axes=F,plg=list(cex=0.75),legend="bottomleft")
sbar(below="m",cex=0.9)
lines(bas)
}

```

We are going to load the corresponding data as dataframes (`x`, `y`, `z`), either from the disk or directly from the github repository (`path_to_data` indicates where to look for the data).

```{r dm5, eval=TRUE, include=FALSE}
path_to_data = "data/gis/"
#path_to_data ="https://raw.githubusercontent.com/VincentGodard/TCNbook/main/data/gis/" # we download directly from the github repository
names_basins = c("Marsyandi","Chudi","Khudi")
cols = c("deepskyblue", "khaki3", "darkolivegreen3")
basins = list() # we store the dataframe into a list object
for (i in 1:length(names_basins)){
  basins[[i]] = read.table(paste0(path_to_data,names_basins[i],".dat"),header=T)
}
```

```{r dm5b, eval=FALSE, include=TRUE}
#path_to_data = "data/gis/"
path_to_data ="https://raw.githubusercontent.com/VincentGodard/TCNbook/main/data/gis/" # we download directly from the github repository
names_basins = c("Marsyandi","Chudi","Khudi")
cols = c("deepskyblue", "khaki3", "darkolivegreen3")
basins = list() # we store the dataframe into a list object
for (i in 1:length(names_basins)){
  basins[[i]] = read.table(paste0(path_to_data,names_basins[i],".dat"),header=T)
}
```

We now plot the distribution of elevation for the 3 basins.



```{r dm6}
#
for (i in 1:length(names_basins)){
  if (i == 1){ # we initiate a plot at the first iteration
    plot(density(basins[[i]]$z),ylim=c(0,0.0022),xlab="Elevation (m)",col=cols[i],main="")
  }else{ # then add the following basins as lines on the initial plot
    lines(density(basins[[i]]$z),col=cols[i])
  }
}
legend("topright",names_basins,cex=0.5,lty=1,col=cols)
```

:::: {.questionbox .questionimage data-latex="question"}
Compare the distributions of elevation, and their position across the Himalayan range.
How is this going to impact the determination of an average production rate for the basin?
::::

We then compute the `st` scaling parameters for each pixel (lines in the dataframes).

```{r dm7}
for (i in 1:length(names_basins)){
basins[[i]]$P = atm_pressure(alt=basins[[i]]$z,model="stone2000") 
st = scaling_st(basins[[i]]$P,basins[[i]]$y)
basins[[i]] = cbind(basins[[i]],st)
}
```

We plot the distribution of the spallation scaling factor for each basin.

```{r dm8}
# plotting result
for (i in 1:length(names_basins)){
  if (i == 1){ # we initiate a plot at the first iteration
    plot(density(basins[[i]]$Nneutrons),ylim=c(1e-4,2.5),xlab="Spallation scaling factor",col=cols[i],main="",log="y")
  }else{
    lines(density(basins[[i]]$Nneutrons),col=cols[i])
  }
}
legend("topright",names_basins,cex=0.5,lty=1,col=cols)
```
We now compute average values for the scaling paramters, in 2 different ways,

- **1** : We compute the scaling at each pixel according to its latitude and altitude (as we did above), and then take the average over the basin
- **2** : we compute the average latitude and altitude of the basin, and then use these average values to compute the scaling factors

We create a dataframe `res` to store the results.

```{r dm9}
res = data.frame(names=names_basins) # dataframe to store the results
res$S1 = NA
res$S2 = NA
res$relief = NA
for (i in 1:nrow(res)){
  res$relief[i] = max(basins[[i]]$z) - min(basins[[i]]$z)
  #
  res$S1[i] = mean(basins[[i]]$Nneutrons)
  #
  P2 = atm_pressure(alt=mean(basins[[i]]$z),model="stone2000") 
  st = scaling_st(P2,mean(basins[[i]]$y))
  res$S2[i] = st$Nneutrons
}
print(res)
```
We can visualize the difference between the two approaches.

```{r dm10}
plot(res$S2,res$S1,pch=21,bg=cols,
     xlab="S2 : Scaling from average elevation",ylab="S1 : Scaling pixel by pixel",
     xlim=range(0,res$S2,res$S1),ylim=range(0,res$S2,res$S1))
grid()
text(res$S2,res$S1,res$names,cex=0.5,pos=1)
abline(0,1,lty=2)

```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce the figures presented in this section
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('dm0','dm5b','dm6','dm7','dm8','dm9','dm10')}
```
::::

:::: {.questionbox .questionimage data-latex="question"}
- Comment on the differences. 
- What are the other parameters which can influence the calculation of average production rates at the scale of a basin?
::::


## Change in denudation through time


## On the importance of muons

For $^{10}$Be the muonic represent a very small fraction of surface production : 

```{r dm20}
nuc = "Be10"
f = (prm[2,nuc]+prm[3,nuc])/sum(prm[1:3,nuc])*100
paste("For",nuc,"muons represent",round(f,1),"% of SLHL surface production")
```

Muons provide a a very small fraction of production at the surface for $^{10}$Be, but they penetrate much deeper than neutrons into rocks. 
For that reason they are going to account to a higher proportion of nuclides on eroding surfaces.
We can separate the various sources of production (neutrons and muons) and see what is their respective contributions to the concentration observed at the surface.


```{r dm21}
col_sp = "deepskyblue"
col_sm = "indianred1"
col_fm = "darkseagreen1"
nuc = "Be10"
data = data.frame(ero1=10^seq(log10(0.1),log10(2000),length=100)) # ero1 -> denudation in m/Ma
emin = min(data$ero1)
emax = max(data$ero1)
data$ero2 = data$ero1/1e6*100*rho # ero2 = denudation in g/cm2/a
# steady state concentrations associated with individual production pathways
data$Csp = prm[1,nuc]*S$Nneutrons/(prm[4,nuc]+(data$ero2/Lambda[1]))
data$Csm = prm[2,nuc]*S$Nmuons/(prm[4,nuc]+(data$ero2/Lambda[2]))
data$Cfm = prm[3,nuc]*S$Nmuons/(prm[4,nuc]+(data$ero2/Lambda[3]))
data$C = data$Csp + data$Csm + data$Cfm
#
plot(NA,xlim=c(emin,emax),ylim=c(0,1),log="x",xlab="Denudation rate (m/Ma)",ylab="Fraction",xaxs="i",yaxs="i")
polygon(c(emin,emax,emax,emin),c(0,0,1,1),col=col_fm)
polygon(c(emin,emax,rev(data$ero1)),
        c(0,0,rev((data$Csp+data$Csm)/data$C)),col=col_sm)
polygon(c(emin,emax,rev(data$ero1)),
        c(0,0,rev(data$Csp/data$C)),col=col_sp)
grid(col="black",equilogs = FALSE)
#
text(0.2,0.1,nuc,cex=2)
legend("bottomright",
       c("Spallation","Stopping muons","Fast muons"),
       pch=22,pt.bg=c(col_sp,col_sm,col_fm),pt.cex=1.5,cex=1,bg="white")
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce the figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('dm0','dm1','dm21')}
```
::::


