# Depth profiles

```{r pr0, include=FALSE}
library("TCNtools")
data(Lambda)
data(prm)
#
```

## Setup and dataset

The package `TCNtools` includes a selection of depth profiles data from the literature. 
We will use the profile published in @laloy2017bayesian from the Campines Plateau in Belgium.

```{r pr1}
data("tcn_depth_profiles") # import the data set
data = tcn_depth_profiles[tcn_depth_profiles$study=="laloy2017bayesian",] # selecting a particular study
data
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(leaflet)
df = data.frame(X =  5.63,Y = 50.95,label="Profile location")
  
m <- leaflet(df) %>%
  addTiles() %>%
  addProviderTiles("OpenStreetMap",group = "OpenStreetMap") %>%
  addProviderTiles("Esri.WorldImagery",group = "Esri.WorldImagery") %>%
  addLayersControl(baseGroups = c("OpenStreetMap","Esri.WorldImagery"),position = "topleft") %>%
  addLabelOnlyMarkers(data = df,
                      lng = ~X, lat = ~Y, label = ~label, 
                      labelOptions = labelOptions(noHide = TRUE, direction = 'auto', textOnly = FALSE)) %>%
  addScaleBar()
m
```


We can then compute the scaling for this study site. 
Note that each site characteristics are only reported for the first sample (`label`=1).

```{r pr2}
altitude = data$altitude[1] # elevation in m
latitude = data$latitude[1] # latitude in degrees
P = atm_pressure(alt=altitude,model="stone2000") # compute atmospheric pressure at site
S = scaling_st(P,latitude) # compute the scaling parameters according to Stone (2000)
```

We plot the depth profile

```{r pr3}
plot(NA,xlim=range(data$C,50e3,200e3),ylim=rev(range(data$depth)),
     xlab="10Be concentration (at/g)",ylab="Depth below surface (cm)")
grid()
arrows(data$C-data$C_e,data$depth,data$C+data$C_e,data$depth,length = 0)
points(data$C,data$depth,pch=21,cex=2,bg="pink")
text(data$C,data$depth,data$label,cex=0.7)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce this figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('pr0','pr1','pr3')}
```
::::

:::: {.questionbox .questionimage data-latex="question"}
Comment on the evolution of concentrations with depth. 
::::

## Profile modelling

We are going to use equation \@ref(eq:conceuler) to compute a theoretical concentration profile.

We use initial guess of the value for the parameters

- Material density $\rho$ (`rho`)
- Surface denudation $\varepsilon$ (`ero`)
- Age of the deposit $t$ (`age`)
- Inherited concentration at the time of deposition $C_0$ (`C0`)

```{r pr4}
rho = 1.7 #  density (g/cm3)
ero = 40*100/1e6*rho  # m/Ma -> g/cm2/a
age = 0.3e6 # a
C0 = 20e3 # inheritance (at/g)
z = seq(0,max(data$depth),length.out = 40)*rho # calculation depth (g/cm2)
C = solv_conc_eul(z,ero,age,C0,prm[,"Be10"],S,Lambda) # compute concentration´
```

We now plot this modeled profile with the dataset.

```{r pr5}
plot(NA,xlim=range(data$C,50e3,200e3),ylim=rev(range(data$depth)),
     xlab="10Be concentration (at/g)",ylab="Depth below surface (cm)")
grid()
lines(C,z,lwd=3) # plotting the modelled profile
arrows(data$C-data$C_e,data$depth,data$C+data$C_e,data$depth,length = 0)
points(data$C,data$depth,pch=21,cex=2,bg="pink")
text(data$C,data$depth,data$label,cex=0.7)
```

:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce this figure
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('pr0','pr1','pr2','pr4',"pr5")}
```
::::

:::: {.actionbox .actionimage data-latex="action"}
Try to get as close to the observations by modifying the various parameters.
::::

:::: {.questionbox .questionimage data-latex="question"}
Comment on the influence of each parameter on the profile.
::::

## Exploration of parameters space

We could change the parameters incrementally one by one to explore the effects on the modeled profile and attempt to converge toward a best-looking solution.
It is obviously much more efficient and robust to do this automatically by going through a large number of parameters sets, compute the predicted concentrations ($C_{mod}$) and evaluate how well they match the measured concentrations ($C_{mes}$ with uncertainty $\sigma$). 
We can use the $\chi^2$ to quantify the difference between the model and observations.

\begin{equation}
\chi^2 = \sum_{i=1}^n \left( \frac{C_{mes,i} - C_{mod,i} }{\sigma_i} \right)^2
\end{equation}

The function `depth_profile_mc` will allow us to generate a large number of models over the parameter space defined by :

- the age of the formation (`age` in a)
- the surface denudation rate (`ero` in m/Ma)
- the inherited concentration (`inh` in at/g)
- the density (`rho` in g/cm2)

There two additional parameters `n1` and `n2`, which control how many models we are going to sample from this parameters space

- `n1` is the number of model which are randomly drawn (Monte Carlo)
- `n2` is the number of regularly sampled points for each parameter (if all 4 parameters are varying the number of models is `n2`$^4$) 

We can choose to carry out an exploration of the parameters space which completely random (`n2=0`) or following a regular grid (`n1=0`), or a combination of both. 

This first attempt is a crude exploration of the parameter space.

```{r pr6}
#data = data[!data$label %in% c(3,5),] # if we want to remove samples 3 and 5 (use with caution)
res = depth_profile_mc(data$C,data$C_e,data$depth,prm[,"Be10"],Lambda,S,
                       age=c(0,100)*1000,
                       ero=c(0,50),
                       inh=c(0,1)*1e5,
                       rho=c(1.5,2.5),
                       n1=50e3)

```

Note that in `depth_profile_mc` we could choose to fix a parameter by passing a single value as argument, instead of a two elements vector, which corresponds to the bounds of the explored interval.

We can have a look at the result table.

```{r}
summary(res)
```
Based on the $\chi2$ value we can now try to have a look at a subset of our model, which correspond to the best fitting values for this metric.
We can start to do that very coarsely by looking at a quantile, for example the best 1%.

```{r pr7}
cutoff = quantile(res$chi2,0.01)
best = res[res$chi2<cutoff,]
pairs(best[,-5],pch=16,col=adjustcolor("blue",0.2)) # we remove the last column (5) which correspond to the chi2
```
We see numerous tradeoffs between parameters, notably between age and denudation rate.

We can select and plot the best model

```{r pr8}
imin = which.min(res$chi2) # position of the lowest chi2 in the dataframe
res[imin,]
z = seq(0,max(data$depth),length.out = 40) # calculation depth (g/cm2)
C = solv_conc_eul(z*res$rho[imin],res$ero[imin]*100/1e6*res$rho[imin],res$age[imin],res$inh[imin],prm[,"Be10"],S,Lambda) # compute concentration
#
plot(NA,xlim=range(data$C,50e3,200e3),ylim=rev(range(data$depth)),
     xlab="10Be concentration (at/g)",ylab="Depth below surface (cm)")
grid()
lines(C,z,lwd=3) # plotting the modeled profile
arrows(data$C-data$C_e,data$depth,data$C+data$C_e,data$depth,length = 0)
points(data$C,data$depth,pch=21,cex=2,bg="pink")
text(data$C,data$depth,data$label,cex=0.7)
```


:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce these figures
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('pr0','pr1','pr2','pr4',"pr6","pr7","pr8")}
```
::::


:::: {.actionbox .actionimage data-latex="action"}
- Change the range of explored parameters
- Try to remove some samples, and observe the influence on the solution
::::


In many situations we have some knowledge about the values or plausible ranges of some parameters. 
For example the density can be measured at the sampling site, and inheritance can be estimated from the shape of the profile. 
In our case it seems reasonable to consider that it is close to 90$\times10^3$ at/g.
We can restrict the range of variation or even fix these parameters, which is what we do next, by letting only the age and denudation rate as free parameters.


```{r pr9}
#data = data[!data$label %in% c(3,5),] # if we want to remove samples 3 and 5 (use with caution)
res = depth_profile_mc(data$C,data$C_e,data$depth,prm[,"Be10"],Lambda,S,
                       age=c(0,200)*1000,
                       ero=c(0,50),
                       inh=90*1e3,
                       rho=2,
                       n1=20000,n2=20)
cutoff = quantile(res$chi2,0.01)
best = res[res$chi2<cutoff,]
plot(best$ero,best$age/1000,pch=16,col=adjustcolor("blue",0.2),
     xlab="Denudation rate (m/Ma)",ylab="Age (ka)")
imin = which.min(res$chi2) # position of the lowest chi2 in the dataframe
points(res[imin,]$ero,res[imin,]$age/1000,col="red")
```

In this case the results will be much easier to visualize through the construction of a $\chi^2$ surface.
We use the library `akima` to grid the results (note that we use $\log_{10}(\chi^2)$).

```{r pr10}
library("akima")
spline<-interp(res$ero/max(res$ero),res$age/max(res$age),log10(res$chi2),duplicate="mean",nx=100,ny=100)
contour(spline$x*max(res$ero),spline$y*max(res$age)/1e3,spline$z,col="pink",xlab="Denudation rate (m/Ma)",ylab="Age (ka)")
points(res[imin,]$ero,res[imin,]$age/1000,col="red") # "best" model
```


:::: {.codebox .codeimage data-latex="code"}
Below you can get all the necessary code to produce these figures
```{r eval=FALSE, class.source='fold-hide', include=TRUE, ref.label=c('pr0','pr1','pr2','pr4',"pr9","pr10")}
```
::::


:::: {.actionbox .actionimage data-latex="action"}
- Change the values for density and inheritance
- Repeat the calculations with the other profiles included in the dataset (`data("tcn_depth_profiles")`) : @siame2004local, @hein2009middle, @hidy2010geologically
::::

:::: {.questionbox .questionimage data-latex="question"}
How well are the age and denudation rate defined for this profile?
::::


## Interactive shiny app

You can also explore profile modelling dynamically  using this embedded application, which use the same type of code. 

```{r echo=FALSE}
knitr::include_url("https://shinyproxy.osupytheas.fr/profile_modelling/",height = "1200px")
```

